Synthesizing Symmetric Lenses

### 2018-06-30 papers/SymmetricLenses/main.tex
in progress edits to section 2

Simple symmetric lenses synchronize files using four functions,
two \emph{create} functions, 
and two \emph{put} functions. Create functions are used to initialize data in
one format when provided with data in the other. If a company.mgmt file is
initialized, a synchronized company.hr file should also be created, and vice
versa. Put functions are used to propagate edits from one format to the other.
If a company.hr file is updated, we should be able to combine it with a
non-updated company.mgmt file, and output a new company.mgmt file, synchronized
to the updated company.hr. To have a simple symmetric lens is to have these
required synchronization functions. Formally, a simple symmetric lens between
$X$ and $Y$ comprises 4 functions:
\begin{enumerate}
\item $\CreateR \OfType X \rightarrow Y$
\item $\CreateL \OfType Y \rightarrow X$
\item $\PutR \OfType X \rightarrow Y \rightarrow Y$
\item $\PutL \OfType Y \rightarrow X \rightarrow X$
\end{enumerate}
Furthermore, if a file is not edited, but is then resynchronized using a put
function, no data should be changed. To be a simple symmetric lens, one must
satisfy the following ``round-tripping'' laws, guaranteeing identity edits won't
change the file data.
\begin{gather}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{x})}{x} = x\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{y})}{y} = y\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{x}{y})}{x} = x\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{y}{x})}{y} = y
\end{gather}

For synchronization, when the original company.hr or company.mgmt file is
created, a synchronized file of the other form is created from it with
$\CreateR$ or $\CreateL$ (depending on which file was initially created). On
future edits, changes are made with the puts, a \PutR would propagate the shared
information from the company.hr file, and knit it together with the information
specific to the company.mgmt file.

\subsection{Programming Symmetric Lenses} We now build up a symmetric lens by
example. Our symmetric lens expressions are typed between regular expression
pairs, if \lstinline{$\Lens$ : $\Regex \Leftrightarrow \RegexAlt$}, then $\Lens$
encodes a simple symmetric lens between $\LanguageOf{\Regex}$ and
$\LanguageOf{\RegexAlt}$. At the end, we want our symmetric lenses to by typed
between \lstinline{emp_salaries} and \lstinline{emp_insurance}.  The finished
lens is shown in Figure~\ref{fig:example_lens}.

The identity lens takes, as argument, a regular expression, and performs the
identity function on strings in that regular expression. The identity lens fully
matches the data formats: when one data format is updated, the other format gets
the exact same value. In our example, when a name is updated, it is also
expected to updated in the other format. The identity lens propagates that
information.
%
\begin{lstlisting}
id(name) : name $\Leftrightarrow$ name
\end{lstlisting}
%

Where the identity lens aligns data, the disconnect lens does the opposite: it
leaves data unaligned. The disconnect lens takes 4 arguments: two regular
expressions and two strings. On creates, because there is no association between
data in the first format and data in the second format, designated default
strings are used for the other format. In our extended example, salaries and
insurance companies should not be matched, so a disconnect lens would be used.
We want our default behavior to specify that the value has not been set.
%
\begin{lstlisting}
disc(salary,company,"UNSET","UNSET") : salary $\Leftrightarrow$ company
\end{lstlisting}
%
When building a disconnect lens between two strings, one of which is merely the
empty string, the lenses \lstinline{ins} and \lstinline{del} can instead be
used, where \lstinline{ins($t$) = disc("",$t$,"",$t$)} and
\lstinline{del($s$) = disc($s$,"",$s$,"")}.

Finally, there are a number of lenses that compositionally build up new lenses
from old ones. For example, data can appear in one form or another: a program
can run during a specific set of minutes, or that program can run the every
minute. If \lstinline{minutes_map} and \lstinline{alltimes_map} are lenses that
synchronize those settings in each format, then those lenses can be composed
with an \OrLens lens.
%
\begin{lstlisting}
or(id(""),multi_emp_map) : emp_salaries $\Leftrightarrow$ "" | ("\n" emp_row) . ("\n" emp_row)*
\end{lstlisting}

Data can also be concatenated, and so lens mappings can be concatenated.  This
is achieved with the concat map.
%
\begin{lstlisting}
concat(ins(header),multi_emp_map) : emp_salaries $\Leftrightarrow$ emp_insurance
\end{lstlisting}
%
Two lenses, $\Lens_1$ and $\Lens_2$, can be concatenated as
\lstinline{concat($\Lens_1$,$\Lens_2$)} or as the infix style
\lstinline{$\Lens_1$ . $\Lens_2$}.

Lastly, the \IterateLens lens synchronizes repeated data, aligning the number of
times the data is repeated. The \IterateLens combinator takes a lens as an
argument and synchronizes each element of the iteration with that lens. For
example, if \lstinline{single_emp_map} synchronizes a single employee's data,
iterating that lens builds a lens that synchronizes a list of tasks.
%
\begin{lstlisting}
iterate(id("\n") . single_emp_map) : ("\n" . emp_salary)$^*$ $\Leftrightarrow$ ("\n" . emp_row)$^*$
\end{lstlisting}
%

\begin{figure}
\begin{lstlisting}
let name_map = id(name) . ins(",") . id(name) . del(": ") . ins(",") in
let single_emp_map = name_map . disc(salary,company,"UNSET","UNSET") in
let multi_emp_map = ins("\n") . single_emp_map . iterate(id("\n") . single_emp_map) in
let emp_map = or(id(""),multi_emp_map) in
ins(header) . multi_emp_map : emp_salaries $\Leftrightarrow$ emp_insurance
\end{lstlisting}
  \caption{A lens, written in our language for building simple symmetric lenses
    on string data, that synchronizes management and hr employee files.
    Note that we can annotate our lenses with types, to confirm the lens
    operates on the languages of the regular expressions. In this work, we
    synthesize the lens itself from such regular expressions.}
  \label{fig:example_lens}
\end{figure}

Using these combinators, we can build up a lens that synchronizes HR and
management employee files. This lens uses one typing rule not previously
mentioned: type equivalence. While \lstinline{scheduled_job} is a equivalent to
a regular expression with an outermost disjunction, it is not syntactically
equivalent to one. Without type equivalence, the provided lens wouldn't actually
be well-typed, but type equivalence states that if a lens is typed between a
pair of regular expressions, it is also typed between an equivalent pair.

Note that, while the programmed lens should be quite simple, it was relatively
difficult to write. We had to provide a lot of scaffolding to map between the
desired formats, even though the idea was quite simple: align the times but
leave the commands unaligned. Furthermore, this is inherent to manually writing
these synchronization functions as a programmer must describe which components
of the strings are aligned with which, and which components are scaffolding.

>>>>

\subsection{Simple Symmetric Lenses}
A simple symmetric lens between formats described by types $\Regex$
and $\RegexAlt$ 
is defined by four functions: 
\begin{enumerate}
\item $\CreateR \OfType \Regex \rightarrow \RegexAlt$
\item $\CreateL \OfType \RegexAlt \rightarrow \Regex$
\item $\PutR \OfType \Regex \rightarrow \RegexAlt \rightarrow \RegexAlt$
\item $\PutL \OfType \RegexAlt \rightarrow \Regex \rightarrow \Regex$
\end{enumerate}
\noindent
that collectively satisfy four ``round-tripping'' laws: 
\begin{gather}
  \tag{\CreatePutRL}
  \PutLOf{(\CreateROf{s})}{s} = s\\
  \tag{\CreatePutLR}
  \PutROf{(\CreateLOf{t})}{t} = t\\
  \tag{\PutRL}
  \PutLOf{(\PutROf{s}{t})}{s} = s\\
  \tag{\PutLR}
  \PutROf{(\PutLOf{t}{s})}{t} = t
\end{gather}
\noindent
The two \emph{create} functions provide default values for times when
data appears on one side but not the other (eg, the ``N/A'' salary
entry added to the management file when HR inserts a new employee).
The two \emph{put} functions propagate edits from one format to the
other by combining the change from one file with the previous value
from the other file. The roundtripping laws guarantee that identify edits to one file do
not change the data in the other file.

%%For synchronization, when the original company.hr or company.mgmt
%%file is created, a synchronized file of the other form is created
%%from it with $\CreateR$ or $\CreateL$ (depending on which file was
%%initially created). On future edits, changes are made with the puts,
%%a \PutR would propagate the shared information from the company.hr
%%file, and knit it together with the information specific to the company.mgmt file.

We use regular expressions to denote the types of data formats, and we
impose a typing discipline on our lenses.  Specifically, if 
$\Regex$ and $\RegexAlt$ are regular expressions, then the notation
\lstinline{$\Lens$ : $\Regex \Leftrightarrow \RegexAlt$} indicates that $\Lens$
is a simple symmetric lens between formats described by $\LanguageOf{\Regex}$ and
$\LanguageOf{\RegexAlt}$. 
When clear from context, we use the term \textit{lens} to denote a
simple symmetric lens.

\subsection{Simple Symmetric Lens Combinators} 
\ksf{This section needs more polishing}
Generally, we do not expect programmers to define simple symmetric lenses by
writing the four functions by hand and showing they satisfy the
roundtripping laws.  Instead, we provide simple symmetric lens
combinators to build up the requisite lenses from simpler ones.
We explain these combinators by using them to define a lens between
the two employee data formats in Figure~\ref{fig:minimized-representations}. 
The finished lens, which appears in Figure~\ref{fig:example_lens}, has
the type 
\lstinline{emp_salaries} $\Leftrightarrow$ \lstinline{emp_insurance}, where
\lstinline{emp_salaries} and \lstinline{emp_insurance} are regular
expressions defined in Figure~\ref{fig:minimized-representations}.

The simplest combinator is the identity lens, which takes as an argument a
regular expression, and performs the 
identity function on strings in that regular expression. The identity lens fully
matches the data formats: when one data format is updated, the other format gets
the exact same value. In our example, when a name is updated, it is also
expected to be updated in the other format. The identity lens propagates that
information.
%
\begin{lstlisting}
id(name) : name $\Leftrightarrow$ name
\end{lstlisting}
%

In contrast to the identity lens, the disconnect lens leaves data
unaligned. The disconnect lens takes four arguments: two regular 
expressions and two strings. The regular expressions specify the
formats of the data on the two sides while the strings provide default
values.  In our extended example, salaries and
insurance companies unrelated, so we use a disconnect lens.  The
defaults are strings that indicate the information is missing.
%
\begin{lstlisting}
disc(salary,company,"N/A","UNK") : salary $\Leftrightarrow$ company
\end{lstlisting}
%
When building a disconnect lens between two strings, one of which is merely the
empty string, the lenses \lstinline{ins} and \lstinline{del} can instead be
used, where \lstinline{ins($t$) = disc("",$t$,"",$t$)} and
\lstinline{del($s$) = disc($s$,"",$s$,"")}.

Finally, there are a number of lenses that compositionally build up new lenses
from old ones. 
For example, data can be concatenated, and so we provide
the \lstinline{concat} combinator that sequences lenses, usually
written using an infix notation \lstinline{$\Lens_1$ . $\Lens_2$}:
%
\begin{lstlisting}
id(name) . ins(",") . id(name) . del(": ") . ins(",")
\end{lstlisting}
% concat(ins(header),employees_lens) : emp_salaries $\Leftrightarrow$ emp_insurance
%

Data can also appear in multiple forms. For example,
a list of employees with insurance information can either be empty or
be a sequence of employee/insurance records: \ksf{This example is
vacuous as you could eliminate the or in favor of the *.  Do somethign
else to introduce or?}
%
\begin{lstlisting}
or(id(""),employees_lens) : emp_salaries $\Leftrightarrow$ "" | ("\n" emp_ins) . ("\n" emp_ins)*
\end{lstlisting}

Lastly, the \IterateLens lens synchronizes repeated data, aligning the number of
times the data is repeated. The \IterateLens combinator takes a lens as an
argument and synchronizes each element of the iteration with that lens. For
example, if \lstinline{employee_lens} synchronizes a single employee's data,
iterating that lens builds a lens that synchronizes a list of employees.
%
\begin{lstlisting}
iterate(id("\n") . employee_lens) : ("\n" . emp_salary)$^*$ $\Leftrightarrow$ ("\n" . emp_ins)$^*$
\end{lstlisting}
%

\begin{figure}
\begin{lstlisting}
let name_lens = id(name) . ins(",") . id(name) . del(": ") . ins(",") in
let employee_lens = name_lens . disc(salary,company,"N/A","UNK") in
let employees_lens = ins("\n") . employee_lens . iterate(id("\n") . employee_lens) in
let emp_lens = or(id(""),employees_lens) in
ins(header) . employees_lens : emp_salaries $\Leftrightarrow$ emp_insurance
\end{lstlisting}
  \caption{A lens, written in our language for building simple symmetric lenses
    on string data, that synchronizes management and HR employee files.
    Note that we can annotate our lenses with types, to confirm the lens
    operates on the languages of the regular expressions. In this work, we
    synthesize the lens itself from such regular expressions.}
  \label{fig:example_lens}
\end{figure}

Using these combinators, we can build up a lens that synchronizes
management and HR employee files.

\ksf{This paragraph needs to be updated, if nothing else, to reflect
the new example. Also, we haven't mentioned any typing rules so the
first sentence doesn't make sense.  Perhaps the whole topic could be
defered to later where we can explain why such a rule is necessary?}
This lens uses one typing rule not previously
mentioned: type equivalence. While \lstinline{scheduled_job} is a equivalent to
a regular expression with an outermost disjunction, it is not syntactically
equivalent to one. Without type equivalence, the provided lens wouldn't actually
be well-typed, but type equivalence states that if a lens is typed between a
pair of regular expressions, it is also typed between an equivalent pair.

Note that, while the synchronization task is quite simple, the
requisite lens is relatively complex. Furthermore, this complexity is
inherent in manually writing simple symmetric lenses because the
lenses must precisely describe the portions of the data that should be
connected and those that should be disconnected.  For disconnected
pieces, the lens must further specifiy appropriate defaults.
